---
title: "Outbreak Detectives: Can You Stop the Spread?"
author: "R-Ladies+ Melbourne"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 9,
  fig.height = 6
)

source("setup/setup.R")
```

# Emergency Alert

Thirty-six students have returned from a school camp.

Several are sick.

Your team has joined the **Global Outbreak Investigation Unit**.

Possible causes include:

- Influenza A
- Bird flu
- COVID-19
- *Staphylococcus aureus* food poisoning
- Group A *Streptococcus* (Strep A)
- Zombie Virus A

Your job is to follow the evidence.

---

# 1. Hospital records have arrived

First, open the patient file.

```{r load-patients}
patients_raw <- load_patients()
```

Take a look.

```{r look-at-data}
head(patients_raw)
```

```{r column-names}
names(patients_raw)
```

### What do you notice?

- What does one row represent?
- What information do we have about each student?
- Can you spot anything messy?

---

# 2. Clean the data

The hospital entered the records very quickly.

For example:

`Yes`, `YES`, `yes` and `Y`

all mean the same thing.

Instead of fixing every value ourselves, we can use a function.

```{r clean-data}
patients <- clean_data(patients_raw)
```

Did it work?

```{r check-cleaning}
count_cases(patients)
```

### Why clean data?

Computers need us to be consistent.  
If `"Yes"` and `"YES"` are written differently, R may think they are different answers.

---

# 3. How many students are sick?

```{r count-cases}
count_cases(patients)
```

Now look at the students with the highest temperatures.

```{r sickest}
show_sickest(patients)
```

### Your turn

What do you notice about their temperatures?

---

# 4. What symptoms are most common?

```{r symptom-table}
symptom_table(patients)
```

Now make a graph.

```{r symptom-plot}
symptoms <- symptom_table(patients)

ggplot(symptoms, aes(
  x = n,
  y = reorder(symptom, n)
)) +
  geom_col(fill = "purple") +
  labs(
    title = "Symptoms reported by sick students",
    x = "Number of students",
    y = "Symptom"
  )
```

### Try editing the graph

Change `"purple"` to another colour and run the code again.

You can also change the words inside `title =`, `x =` and `y =`.

### Think like a disease detective

Which symptoms stand out?

Does this help rule out any diseases?

---

# 5. Do sick students have higher temperatures?

```{r temperature}
ggplot(patients, aes(
  x = case_status,
  y = temperature_c,
  fill = case_status
)) +
  geom_boxplot() +
  labs(
    title = "Do sick students have higher temperatures?",
    x = "Student group",
    y = "Temperature (°C)"
  )
```

### Try editing the graph

Inside `aes()`, change:

```r
fill = case_status
```

to:

```r
colour = case_status
```

What changes?

### What do you notice?

Compare the middle of the two boxplots.

---

# 6. How quickly did students become sick?

```{r onset}
show_onset(patients)
```

The time between being exposed and becoming sick can be an important clue.

---

# 7. New evidence: locations

A second file has arrived.

```{r locations}
locations <- load_locations()
```

We need information from **both files**.

```{r join}
investigation <- combine_evidence(
  patients,
  locations
)
```

Now find the locations with the most illness.

```{r location-table}
location_table(investigation)
```

And plot them.

```{r location-plot}
locations_summary <- location_table(investigation)

ggplot(locations_summary, aes(
  x = attack_rate,
  y = reorder(location, attack_rate)
)) +
  geom_col(fill = "skyblue") +
  labs(
    title = "Where were sick students exposed?",
    x = "Attack rate",
    y = "Location"
  )
```

### Try editing the graph

Change `"skyblue"` to `"orange"`, `"green"` or another colour.

Then try changing the graph title.

### What does this tell us?

Does one location or activity stand out?

Could there be more than one explanation?

---

# 8. Make a preliminary diagnosis

Open the disease clue sheet.

```{r disease-clues}
disease_clues <- load_disease_clues()

disease_clues
```

Before you see the laboratory results, decide:

1. What disease do you think is spreading?
2. What evidence supports your answer?
3. Is there anything unusual about two of the sick students?

---

# 9. Breaking News: sequencing results

The laboratory has sequenced DNA from several student samples.

We can use a phylogenetic tree to compare them with reference samples.

The difficult code is hidden inside our function.

```{r tree, fig.width=10, fig.height=7}
show_tree(
  sample_colour = "red",
  reference_colour = "grey30",
  title = "Which reference is closest to the outbreak samples?"
)
```

### Read the tree

- Which reference is closest to most of the student samples?
- Do all of the student samples belong to the same group?
- What do you notice about P29 and P30?

A tree can tell us which samples are genetically similar.

It **cannot** prove exactly who infected whom.

---

# 10. Contact tracing

Interviewers have now told us who spent time together.

```{r contacts}
contacts <- load_contacts()
```

How many separate groups are in the contact data?

```{r network-groups}
network_groups(contacts)
```

Now draw the network.

```{r network, fig.width=10, fig.height=7}
network <- show_network(
  contacts,
  patients,
  sick_colour = "red",
  well_colour = "blue",
  title = "Who had contact with whom?"
)

network
```

### Investigate the network

- Are most sick students connected?
- What is unusual about P29 and P30?
- Why might some connected students still be well?
- Does a connection prove that infection occurred?

---

# 11. Final diagnosis

Your team must now give the outbreak response team its answer.

### Main outbreak

What is the most likely disease?

### The two unusual students

Are P29 and P30 part of the same outbreak?

### Give three pieces of evidence

You could use:

- symptoms
- time to illness
- locations
- sequencing
- contact network

---

# The bigger lesson

Real outbreak investigations combine different types of evidence.

No single graph gives us the whole answer.

Today you used:

**data + coding + epidemiology + genomics + networks**

to solve an outbreak.

