---
title: "Outbreak Detectives: Can You Stop the Spread?"
author: "R-Ladies+ Melbourne"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

# Facilitator overview

This 70-minute workshop updates the original **It Takes a Spark** disease-detective activity. It retains the approachable progression from inspecting columns to selecting, filtering, counting, summarising, reshaping and plotting data, but each skill now unlocks a new part of the investigation.

Six mystery folders are provided. Teams follow the same R code but may reach different diagnoses.

> **Scientific note:** All patient records and sequences are simulated. Disease patterns are simplified for teaching. This activity is not a diagnostic tool.

# Suggested timing

| Time | Investigation update | Main R skill |
|---:|---|---|
| 0–7 min | Emergency alert and R orientation | objects, scripts, running code |
| 7–16 min | Hospital records arrive | `read_csv()`, `colnames()`, `glimpse()` |
| 16–27 min | Clean the rushed records | `mutate()`, `case_when()`, text cleaning |
| 27–40 min | Epidemiologist requests symptom summary | `filter()`, `count()`, `summarise()`, `pivot_longer()` |
| 40–50 min | Location data arrive | `left_join()`, grouped summaries, plot |
| 50–56 min | Teams present preliminary findings | interpretation |
| 56–64 min | Sequencing results arrive | phylogenetic-tree reveal |
| 64–68 min | Contact network | network interpretation |
| 68–70 min | Final diagnosis | evidence-based conclusion |

# 🚨 Emergency Alert: several students are sick

Read `updates/00_EMERGENCY_ALERT.txt` aloud.

You are investigators in the **Global Outbreak Investigation Unit**. Thirty-six students have returned from a school science camp, and an unusual cluster of illness has been reported.

Possible causes are:

- Influenza A
- Bird flu
- COVID-19
- *Staphylococcus aureus* food poisoning
- Ebola
- Zombie Virus A

Do not guess yet. Follow the data.

## A very short R warm-up

The original workshop introduced R as both a calculator and a notebook. Retain one quick example so beginners practise running a line of code and saving a result.

```{r warm-up}
2 + 2

team_name <- "The Data Detectives"
paste("Welcome,", team_name)
```

# 📁 The hospital has sent patient records

The facilitator assigns each team a mystery letter. Change `A` below.

```{r choose-case, eval=FALSE}
mystery <- "A"
case_folder <- file.path("case_files", paste0("mystery_", mystery))
```

Load the first evidence file.

```{r load-patients, eval=FALSE}
patients_raw <- read_csv(
  file.path(case_folder, "01_patient_records_messy.csv"),
  show_col_types = FALSE
)
```

## Why are we doing this?

Outbreak investigators first need to understand what information they have. Looking at the data before analysing it prevents incorrect assumptions.

```{r inspect, eval=FALSE}
colnames(patients_raw)
glimpse(patients_raw)
head(patients_raw)
```

Questions:

1. What does one row represent?
2. Which columns contain numbers?
3. Which columns contain categories or words?
4. Can you spot inconsistent values?

# 🧹 The data are messy—clean them first

The records were entered quickly by several hospital staff. “Yes”, “YES”, “yes” and “Y” all mean the same thing.

## Why are we doing this?

Real data are rarely ready for analysis. Computers treat differently written values as different categories unless we standardise them.

```{r clean-data, eval=FALSE}
yes_no_columns <- c(
  "hospitalised", "fever", "cough", "sore_throat", "headache",
  "muscle_aches", "vomiting", "diarrhoea", "loss_of_smell",
  "bleeding", "confusion"
)

patients <- patients_raw %>%
  mutate(
    sex = case_when(
      str_to_lower(sex) %in% c("f", "female") ~ "Female",
      str_to_lower(sex) %in% c("m", "male") ~ "Male",
      TRUE ~ "Another identity / not recorded"
    ),
    case_status = if_else(
      str_to_lower(case_status) %in% c("sick", "yes"),
      "Sick", "Well"
    ),
    across(
      all_of(yes_no_columns),
      ~ if_else(str_to_lower(.x) %in% c("yes", "y"), "Yes", "No")
    )
  )
```

Check whether cleaning worked.

```{r check-cleaning, eval=FALSE}
patients %>% count(sex)
patients %>% count(case_status)
patients %>% count(fever)
```

# 📊 The Chief Epidemiologist wants a symptom summary

## How many students are sick?

```{r count-cases, eval=FALSE}
patients %>% count(case_status)
```

## Who may need urgent attention?

```{r severe-cases, eval=FALSE}
patients %>%
  filter(case_status == "Sick") %>%
  arrange(desc(temperature_c)) %>%
  select(patient_id, first_name, temperature_c, hospitalised)
```

## Why are we doing this?

Filtering and sorting help response teams identify severe cases and decide where medical attention may be needed first.

## Bring all symptoms together

The original workshop introduced the important idea that data can be stored in a wide or long format. Here, each symptom currently has its own column. We reshape the data so every symptom can be counted in one analysis.

```{r reshape-symptoms, eval=FALSE}
symptoms_long <- patients %>%
  filter(case_status == "Sick") %>%
  pivot_longer(
    cols = all_of(setdiff(yes_no_columns, "hospitalised")),
    names_to = "symptom",
    values_to = "present"
  )
```

```{r symptom-summary, eval=FALSE}
symptom_summary <- symptoms_long %>%
  filter(present == "Yes") %>%
  count(symptom, sort = TRUE)

symptom_summary
```

## Visualise the symptom profile

```{r symptom-plot, eval=FALSE}
symptom_summary %>%
  ggplot(aes(x = n, y = fct_reorder(symptom, n))) +
  geom_col() +
  labs(
    title = "Symptoms reported by sick students",
    x = "Number of students",
    y = NULL
  )
```

## Why are we doing this?

Diseases produce different patterns of symptoms. A summary does not prove the cause, but it helps narrow the list of suspects.

## Explore temperature and onset time

The original activity used boxplots to compare temperature across groups. Here the same skill answers a more direct outbreak question.

```{r temperature-plot, eval=FALSE}
patients %>%
  ggplot(aes(x = case_status, y = temperature_c, fill = case_status)) +
  geom_boxplot(show.legend = FALSE) +
  labs(
    title = "Do sick students have higher temperatures?",
    x = NULL,
    y = "Temperature (°C)"
  )
```

```{r onset-summary, eval=FALSE}
patients %>%
  filter(case_status == "Sick") %>%
  summarise(
    cases = n(),
    median_temperature = median(temperature_c, na.rm = TRUE),
    median_hours_to_symptoms = median(hours_to_symptoms, na.rm = TRUE),
    earliest_onset = min(hours_to_symptoms, na.rm = TRUE),
    latest_onset = max(hours_to_symptoms, na.rm = TRUE)
  )
```

# 🗺️ A second file has arrived containing locations

```{r load-locations, eval=FALSE}
locations <- read_csv(
  file.path(case_folder, "02_locations.csv"),
  show_col_types = FALSE
)
```

Join the location information to the cleaned patient records.

```{r join-data, eval=FALSE}
investigation <- patients %>%
  left_join(locations, by = "patient_id")
```

## Why are we doing this?

Useful clues are often spread across separate files. Joining them allows investigators to connect symptoms with possible exposures.

```{r location-summary, eval=FALSE}
location_summary <- investigation %>%
  group_by(location) %>%
  summarise(
    students = n(),
    sick = sum(case_status == "Sick"),
    attack_rate = sick / students,
    .groups = "drop"
  ) %>%
  arrange(desc(attack_rate), desc(sick))

location_summary
```

```{r location-plot, eval=FALSE}
location_summary %>%
  ggplot(aes(x = attack_rate, y = fct_reorder(location, attack_rate))) +
  geom_col() +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title = "Which location has the highest proportion of illness?",
    x = "Attack rate",
    y = NULL
  )
```

> **Teaching point:** An attack rate is the proportion of exposed people who became ill. A high attack rate identifies an important clue, but association alone does not prove causation.

# 📈 Present preliminary findings

Each team has 60 seconds to report:

1. The most common symptoms.
2. The timing of symptom onset.
3. The location or exposure with the highest attack rate.
4. Their preliminary diagnosis.
5. One uncertainty.

Compare your findings with `disease_reference.csv`.

```{r disease-reference, eval=FALSE}
disease_reference <- read_csv("disease_reference.csv", show_col_types = FALSE)
disease_reference
```

# 🧬 Breaking News: genome sequencing results are back

The reference laboratory has sequenced samples from eight patients. Students should first make a prediction using the epidemiological evidence. The tree then acts as an independent laboratory clue.

## Facilitator phylogenetic-tree code

This section is more advanced and can be removed from the student version. It creates a neighbour-joining tree from the simulated aligned DNA sequences, uses a TN93 genetic-distance model, roots the tree at its midpoint and calculates bootstrap support.

Required packages:

```{r phylo-packages, eval=FALSE}
install.packages(c("ape", "phangorn"))
```

```{r phylogenetic-tree, eval=FALSE}
library(ape)
library(phangorn)

alignment_file <- file.path(
  case_folder,
  "03_sequencing_alignment.fasta"
)

dna <- read.dna(
  alignment_file,
  format = "fasta"
)

# Estimate pairwise genetic distances using a nucleotide substitution model.
genetic_distance <- dist.dna(
  dna,
  model = "TN93",
  pairwise.deletion = TRUE
)

# Construct a neighbour-joining tree.
tree_nj <- nj(genetic_distance)

# Midpoint rooting improves presentation when no true outgroup is supplied.
tree_rooted <- midpoint(tree_nj)

# Bootstrap the alignment columns to assess support for major groupings.
set.seed(2026)
bootstrap_support <- boot.phylo(
  phy = tree_rooted,
  x = dna,
  FUN = function(x) {
    midpoint(nj(dist.dna(x, model = "TN93")))
  },
  B = 200,
  quiet = TRUE
)

tip_group <- ifelse(
  grepl("^REF_", tree_rooted$tip.label),
  "Reference pathogen",
  "Outbreak sample"
)

tip_colours <- ifelse(
  tip_group == "Outbreak sample",
  "#88398A",
  "grey30"
)

plot(
  tree_rooted,
  type = "phylogram",
  direction = "rightwards",
  tip.color = tip_colours,
  cex = 0.75,
  label.offset = 0.002,
  main = "Genetic relationship between outbreak and reference samples"
)

# Display bootstrap values of at least 70%.
nodelabels(
  text = ifelse(bootstrap_support >= 70, bootstrap_support, ""),
  frame = "none",
  cex = 0.65,
  adj = c(1.2, -0.2)
)

legend(
  "topleft",
  legend = c("Outbreak sample", "Reference pathogen"),
  col = c("#88398A", "grey30"),
  pch = 19,
  bty = "n"
)
```

## How students interpret the tree

Ask them to identify:

- Which reference tip sits in the same supported cluster as the outbreak samples?
- Do all outbreak samples form one tight group?
- Are any outbreak samples unusually distant from the others?
- Which branches have bootstrap values of 70% or more?

## Facilitator interpretation

The outbreak samples should form a tight cluster with the reference sequence corresponding to the assigned mystery. Short distances among outbreak samples indicate that their simulated sequences are very similar. This supports a shared outbreak rather than unrelated illnesses.

Bootstrap values describe how consistently a grouping appears when alignment positions are resampled. Values above approximately 70% can be presented as reasonably strong support for this classroom exercise, although interpretation depends on the dataset and method.

The tree **does not prove**:

- who infected whom;
- the exact direction of transmission;
- the precise place of infection;
- that every epidemiologically linked patient belongs to the same transmission chain.

Phylogenetic results should therefore be interpreted together with symptoms, onset times, exposures and contact information.

# 🌳 Interpret the phylogenetic tree

Students write one sentence:

> “The outbreak samples are most genetically similar to __________ because __________.”

Then ask whether the tree confirms or changes their preliminary diagnosis.

# 🕸️ Can you reconstruct how the disease spread?

Load the contact data.

```{r contact-data, eval=FALSE}
contacts <- read_csv(
  file.path(case_folder, "04_contacts.csv"),
  show_col_types = FALSE
)
```

Optional facilitator network:

```{r contact-network, eval=FALSE}
install.packages("igraph")
library(igraph)

network <- graph_from_data_frame(
  contacts,
  directed = FALSE,
  vertices = patients %>% select(patient_id, case_status)
)

V(network)$color <- ifelse(
  V(network)$case_status == "Sick",
  "#88398A",
  "grey80"
)

V(network)$size <- ifelse(
  V(network)$case_status == "Sick",
  9,
  6
)

set.seed(2026)
plot(
  network,
  vertex.label.cex = 0.65,
  vertex.frame.color = NA,
  edge.color = "grey70",
  main = "Reported contacts among students"
)
```

## Why are we doing this?

A contact network shows who interacted with whom and can suggest plausible routes of spread. However, a connection does not prove transmission, and missing contacts can change the apparent network.

Discuss:

- Are sick students concentrated in one connected part of the network?
- Who might be an early or highly connected case?
- What extra information would you need before claiming that one person infected another?

# 🏆 Submit your final diagnosis

Teams submit:

1. The most likely diagnosis.
2. Three pieces of supporting evidence:
   - symptom profile;
   - onset timing or exposure;
   - phylogenetic clustering.
3. The likely source or exposure.
4. One limitation.
5. One recommended public-health action.

# Facilitator answer key

```{r answer-key, echo=FALSE}
read_csv("facilitator/answer_key.csv", show_col_types = FALSE)
```

# What was repurposed from the original workshop?

The revised activity retains:

- a very short calculator/notebook introduction;
- inspecting column names and data types;
- selecting columns;
- plotting temperature;
- using boxplots;
- generating grouped summaries;
- counting symptoms;
- reshaping symptom data;
- filtering to sick students;
- adding a second location dataset.

The main change is that these are no longer isolated exercises. Each command answers a question requested by the outbreak-response team.
